# The impact of the career length of directors on their IMDb ratings

----
title: "The impact of the career length of directors on their IMDb ratings"
author: "Joël de Vries, Edwin van den Dikkenberg, Paulien Beeker"
date: "2025-03-22"
output: pdf_document
----

```{r include=FALSE}
## Installing and loading necessary packages
if (!requireNamespace("tinytex", quietly = TRUE)) install.packages("tinytex")
library(tinytex)
if (!requireNamespace("knitr", quietly = TRUE)) install.packages("knitr")
library(knitr)
```


## Introduction

The film industry is shaped by the talent and vision of directors, whose creative decisions significantly impact the quality and reception of their movies. This study explores the relationship between a director’s career length and the ratings their films receive on IMDb, providing insight into whether experience and longevity in filmmaking translate into higher audience appreciation.

## Motivation

This study is relevant as it can provide insights into whether experience plays a significant role in filmmaking success, informing both aspiring and established directors, film producers, and scholars studying the film industry

By analyzing IMDb datasets, we will measure a director’s career length based on the span between their first and last directed movie and compare this to the average IMDb ratings of their films. Additionally, the number of productions attributed to each director will be taken into account, allowing us to explore whether a higher volume of work influences overall ratings.

Furthermore, genre versatility—the number of different genre categories a director has worked in—will be considered as a control variable. Directors who have experience across multiple genres may demonstrate greater adaptability and creative range, which could positively influence audience appreciation and ratings. On the other hand, directors specializing in specific genres might develop a strong niche following, potentially leading to higher ratings within their expertise.

The findings could provide insight into whether experience, productivity, and creative diversity correlate with higher audience appreciation.

## Data

The research was conducted based on the following datasets extracted from IMDb

-  `"title.basics"` = <https://datasets.imdbws.com/title.basics.tsv.gz>  
-  `"title.crew"` = <https://datasets.imdbws.com/title.crew.tsv.gz>  
-  `"title.ratings"` = <https://datasets.imdbws.com/title.ratings.tsv.gz>  
-  `"name.basics"` = <https://datasets.imdbws.com/name.basics.tsv.gz>

Further exploration of the original dataset can be found in the following document:
- "gen/output/data.exploration"

### Variable description

| Variable name     | Variable explanation | 
|--------------|--------------------|
| career_start  | year of production of the first movie directed                 | 
| career_end | year of production of the last movie directed                 | 
| career_length  |   total length of a director's career, calculated as career_end - career_start               | 
| num_movies  |  total amount of movies produced per director  |
| avg_runtime  |  average runtime of movies per director  |
| avg_numVotes  |  average number of votes for movies per director  |
| avg_rating  |  average rating of movies produced rated from 0-10  |
| career_category |  short (<10 years), medium (10-30 years), long (>30 years) |
|  is_dramatic |  indicates if the movie belongs to the dramatic genres drama, biography, history (1=yes, 2=no)  |
|  is_light_action   |  indicates if the movie belongs to the light entertainment genres comedy, musical, music (1=yes, 2=no) |
|  is_action_suspense   |  indicates if the movie belongs to action or suspense genres action, adverture, thriller, crime (1=yes, 2=no)   |
|  is_speculative |  indicates if the movie belongs to speculative genres horror, mystery, sci-fi, fantasy (1=yes, 2=no)   |
|  is_non_fiction |  indicates if the movie belongs to non-fiction genres documentary, news, reality-TV, talkshow (1=yes, 2=no)  |
|  genre_versatality |  amount of genre categories a director has worked on   |

## Methodology

To answer this question, a multiple regression analysis is chosen as the primary research method. Regression is well-suited for this study because it allows us to examine the relationship between a director’s career length and the average IMDb rating of their movies while also considering the number of productions they have directed.

Additionally, genre versatility is included in the regression model to account for the possibility that working across multiple genres influences a director’s ability to maintain high ratings.

The regression model is as follows: Y = avg_rating ~ career_length + num_movies + avg_runtime + avg_numVotes + is_dramatic + is_light_entertainment + is_action_suspense + is_speculative + is_non_fiction + genre_versatility

## Analysis
### Box plot career length categories
Before the multiple regression analysis was run, a box plot was computed to show the differences between a short (<10 years), medium (10-30 years), and long (>30 years) director career on the effect of the IMDb rating.
```{r, echo=FALSE, out.width="80%"}
knitr::include_graphics("../gen/output/box_plot.pdf")
```

### Box plot genre categories

To further investigate the effect of the controle variable genre category, another box plot was created to give insight into the effect of genre category on IMDb ratings.
```{r, echo=FALSE, out.width="80%"}
knitr::include_graphics("../gen/output/genre_category_boxplot.pdf")
```

### Box plot genre versatility

To show the effect of genre versatility (how many different genres) of a director on the IMDb ratings, another box plot was computed. 
```{r, echo=FALSE, out.width="80%"}
knitr::include_graphics("../gen/output/genre_versatility_plot.pdf")
```

### Scatter plot productivity

To get more insight into the effect of the productivity of a director on the film rating, a scatter plot was created. This shows relation between the number of movies that a director direct and the average IMDb rating of all the movies of that particular director.
```{r, echo=FALSE, out.width="70%"}
knitr::include_graphics("../gen/output/productivity_vs_rating.pdf")
```

### Multiple regression analysis

Finally, the multiple regression model, including the independent variable "career length" and all the control variables was run (see Methodology). The regression analysis resulted in the output displayed below.
```{r, echo=FALSE, out.width="80%"}
knitr::include_graphics("../gen/output/model_summary.pdf")
```

## Findings

The analysis shows that career length has a small but significant negative impact on IMDb ratings. Directors with longer careers tend to see a decline in their average ratings. Similarly, directors who produce a high number of films tend to have lower average ratings, suggesting that prioritizing quantity over quality may lead to diminishing returns.

Genre categories significantly influence ratings. Non-fiction and dramatic films tend to receive higher scores, while speculative and action/suspense genres generally score lower. However, genre versatility does not appear to have a significant effect on ratings, indicating that working across multiple genres does not necessarily improve or reduce audience appreciation.

The multiple regression analysis confirms these trends, showing that career length, number of movies directed, and genre category are key factors influencing IMDb ratings. The regression results indicate that career length and the number of movies produced significantly affect average IMDb ratings. A negative trend can be observed for both variables, with the effect being stronger for the number of movies produced than for career length.


## Conclusion

The regression findings provide valuable insights for the film industry, guiding decisions in talent selection, budget allocation, and marketing strategies. For professionals such as producers and studio executives, the data suggests that directors with longer careers may produce films with declining ratings over time. This could prompt studios to focus on directors with shorter, more consistent track records and allocate budgets to those prioritizing quality over quantity. Marketing efforts could also emphasize high-rated works rather than an entire career.

## Recommendations

For streaming platforms like Netflix and Amazon Prime, the results are important for selecting content and adjusting recommendation algorithms. Platforms may prefer directors with a consistent record of quality work and could focus on emerging filmmakers rather than established ones with declining ratings. This strategy helps maintain a balance between new talent and well-known directors.

Academically, the study opens avenues for research on the impact of career longevity on creative output, exploring whether long careers contribute positively or negatively to film quality. It also invites comparisons with other creative industries like music and literature, highlighting potential industry-specific trends and the influence of genre demands on a director's performance.

